---
title: "Week 4 Notes"
output: html_notebook
---

# str() Function
* Compactly display the internal structure of an R object.
- A diagnostic function and an alternative to "summary"
- It is especially well-suited to compactly display the (abbreviated) contents of 
  (possibly nested) lists.
- Roughly one line per basic object.

```{r}
str(str) # str() gives you the arguments that a function takes.
str(lm)
x <- rnorm(100,2,4)
summary(x) # summary() gives statistics BUT...
str(x) # str() gives you more info, like whether or not this is a numeric vector

f <- gl(40,10) # A factor with forty levels
summary(f) # Gives you a table of how many of each level there is in f
str(f) # Gives you the object class and how many levels

library(datasets)
head(airquality) 
str(airquality) # Gives each column and what kind of classes it contains.
m <- matrix(rnorm(100),10,10)
str(m) # Gives us dimensions of matrix
m[,1]

s <- split(airquality, airquality$Month) # Using split to create list separated by levels
str(s) # Using str() on s gives a highly detailed break-down of the list created in split()
```

# Simulation: Generating Random Numbers
* Functions for probability distributions in R
- rnorm: Generate random Normal variates with a given mean and SD
- dnorm: Evaluate Normal probability density (with a given mean/SD) at point (or vector of
  points)
- pnorm: Evaluate the cumulative distribution function for a Normal distribution
- rpois: Generate random Poisson variates with a given rate

* Probability distribution functions usually have four functions associated with them.
* They are prefixed with a
        - d for density
        - r for random number generation
        - p for cumulative distribution
        - q for quantile function
```{r}
str(dnorm) # See what arguments these take
str(pnorm)
str(qnorm)
str(rnorm)
```

* Anytime you generate random numbers, you must set a seed, which will ensure your work
  is reproducible.
```{r}
set.seed(1)
rnorm(5)
# The seed is now unset and must be reset again.
set.seed(1)
rnorm(5) # And you get the same numbers.
```

* Generating poisson data
```{r}
str(rpois)
rpois(10,1)
ppois(2,2) # Cumulative distribution, output is Pr(x <= 2)

```

