---
title: "Week 3 Notes"
output: html_notebook
---
# Loop Functions
* For & while loops are useful when programming but not when working interactively on the 
  command line; there are some functions which implement looping to make life easier
  * lapply: Loop over a list and evaluate function on each element
  * sapply: Same as lapply but simplify the result
  * apply: Apply a function over subsets of a vector
  * tapply: Apply function over vector subsets
  * mapply: Multivariate version of lapply
* Aux function split() also useful in conjuction with lapply

# lapply
- Takes three arguments: a list, a function, and other arguments (...)
- If x is not a list, it will be coerced using as.list
- Always returns a list, regardless of the class of the input
```{r}
# Easy example
a <- list(b=1:5,c=rnorm(10))
a
lapply(a,mean) # Gives mean of first and second elements

# Slightly more complex example
d <- list(e=1:4,f=rnorm(10),g=rnorm(20,1),h=rnorm(100,5))
lapply(d,mean)

# Another way to call lapply
i <- 1:4
lapply(i,runif) # Uniform random variables
lapply(i,runif, min=0,max=10) # Passing arguments through ...
```
* lapply and friends make heavy use of anonymous functions (functions that don't have 
  names)
```{r}
j <- list(a=matrix(1:4,2,2),b=matrix(1:6,3,2))
j # Say you want to extract the first column from these matrices.
lapply(j,function(elt) elt[,1]) # "elt" is an anonymous function!
```

# sapply
- Tries to simplify the result of lapply whenever possible
- The result is a list where every element is length 1, then a vector is returned; 
  where every element is a vector of the same length (>1), a matrix is returned.
- If it can't figure it out, a list is returned
```{r}
k <- list(l = 1:4, m = rnorm(10), n = rnorm(20,1), o = rnorm(100,5))
lapply(k,mean) # Returns a list, but what if we just want a vector?
sapply(k,mean) # sapply does just that.
mean(k) # DOESN'T WORK because k is a list.
```

# apply
- Used to evaluate a function over the margins of an array
- Most often used to apply a function to the rows or columns of a matrix
- Can be used with general array (like taking the average of an array of matrices)
- Not really faster than writing a loop but works in one line!
```{r}
str(apply)
# X is an array
# MARGIN is an integer vector indicating which margins should be "retained"
## THIS CORRESPONDS TO THE DIMENSION. For example, (row column) is (1 2).
# FUN is a function to be applied
# ... for other arguments passing through

# Let's try an example
l <- matrix(rnorm(200), 20, 10) # Create random normal matrix of 200 points
apply(l, 2, mean) # Returns vector of length 10 from each averaged column of matrix l
apply(l, 1, sum) # Returns vector of length 20 from each summed ROW of matrix l
```
* Shortcuts: col/row sums and means; faster if using large matrix
- rowSums = apply(x,1,sum)
- rowMeans = apply(x,1,mean)
- colSums = apply(x,2,sum)
- colMeans = apply(x,2,mean)
* Other ways to apply apply()
- Quantiles of matrix rows
```{r}
apply(l,1,quantile,probs = c(0.25,0.75))
# Creates 2x20 matrix of 25% and 75% quantiles of each row
```
- Average matrix in an array
```{r}
m <- array(rnorm(2*2*10),c(2,2,10)) # Returns 10 matrices of 4 normal random variables
apply(m,c(1,2),mean) # applies mean for each matrices and returns 2x2 matrix
# This is a very important example because, for the MARGIN argument, we have neglected
# the third dimension, meaning that apply will "stack" the 2x2 matrices and then
# collapse them into one, giving the average of each "stack" quadrant. Very cool!
rowMeans(m,dims=2) # Does the same thing as the above.
```

# mapply
* A multivariate apply that applies a function in parallel over a set of arguments
* In other words, you specify a function and apply it to multiple lists of objects
```{r}
str(mapply)
# FUN = Function to apply
# ... = Contains arguments
# MoreArgs = List of other arguments to FUN
# SIMPLIFY = Indicates whether the result should be simplified

# Example
list(rep(1,4), rep(2,3), rep(3,2), rep(4,1))
# Instead of this, we can do:
mapply(rep, 1:4, 4:1)
```

## Vectorizing a Function
```{r}
noise <- function(n, mean, sd) {
        rnorm(n, mean, sd)
}
noise(5, 1, 2)
noise(1:5, 1:5, 2)
mapply(noise,1:5,1:5,2) # Instant vectorization!

# The above mapply() is the same as:
list(noise(1,1,2),noise(2,2,2),noise(3,3,2),noise(4,4,2),noise(5,5,2))
```

# tapply
* Used to apply a function over subsets of a vector.
```{r}
str(tapply)
# X = Vector
# Index = Factor or list of factors
# Fun = Function to apply
# ... = Arguments to be passed
# simplify = Should we simplify result?

# Lets do an example where we take group means
x <- c(rnorm(10), runif(10), rnorm(10,1))
f <- gl(3,10)
f
x
tapply(x, f, mean) # Takes means of each FACTOR (1,2,3) in subsets of x.
# In other words, your first mean will be equivalent to:
mean(x[1:10])
# In the above tapply(), the result was a vector because the default is simplification.
tapply(x,f,mean,simplify=FALSE)
# Setting simplify=FALSE gives a LIST instead.
tapply(x,f,range)
# Gives a list of min and max for each FACTORED GROUP.
```

# split
* Takes a vector or other objects and splits it into groups determined by a factor or
  list of factors.
```{r}
str(split)
# x = Vector or List
# f = factor (or coerced to one) or a list of factors
# drop = indicates whether empty factor levels should be dropped
# From here, you can use lapply or sapply on your curated groups.

# Example:
x <- c(rnorm(10), runif(10), rnorm(10,1)) # Generate list of data
f <- gl(3,10) # Generate list of factors
split(x,f) # SPLIT them according to their levels in f. Indices will be paired up.
# From two vectors, you have created ONE list
# Now we can put our LIST into lapply
lapply(split(x,f),mean)
# Though this was not entirely necessary (we could have done all this with tapply)
# its a decent example of the use of lapply and split() in tandem

# Lets try an example that is more relevant
```

## Splitting a Data Frame
```{r}
library(datasets)
head(airquality) # 100+ rows and 6 columns

# Lets calculate the mean ozone, solar, and wind measurements in each month
# First, we can split the months by level
s <- split(airquality, airquality$Month) # A list of dataframes

# Next, we use lapply to take the column means of each column we are interested in
lapply(s, function(x) colMeans(x[, c("Ozone","Solar.R","Wind")]))
# This gives us a list of vectors.
# We can do the same thing using sapply.
sapply(s, function(x) colMeans(x[, c("Ozone","Solar.R", "Wind")]))
# This gives us a clean 3x5 matrix.
# Note that some means are NA. Let's fix that.
sapply(s, function(x) colMeans(x[, c("Ozone","Solar.R", "Wind")], na.rm=TRUE))
# Add the na.rm=TRUE argument to colMeans excludes NAs from the mean calculation.
```

## Splitting on More than One Level
```{r}
x <- rnorm(10)
f1 <- gl(2,5) # Let's call this factor delimitation "race"
f2 <- gl(5,2) # This one "income"

# What should we do when we want to combine these factors?
interaction(f1,f2) # Combines levels of factor lists f1 and f2
# Lets look at what is the new list composed of with respect to x
str(split(x,list(f1,f2))) # This shows that some levels don't have anything
# How can we drop these vectors?
str(split(x,list(f1,f2),drop=TRUE)) # Like so. 
# Now, only levels with observations in them will be saved.
```

# Debugging Tools
## Diagnosing the Problem
* Indications that something's not right
- message: Generic notification/diagnostic message produced by message function;
  execution of the function continues
- warning: An indication that something is wrong but not necessarily fatal;   
  execution of the function continues; gnerated by the warning() function
- error: An indication that a fatal problem has occured; execution stop;   
  produced by the stop() function
- condition: A generic concept for indicating that something unexpected can 
  occur; programmers can create their own conditions
```{r}
log(-1) # Warning example: NaN produced

# How do we fix/sidestep something like this?
printmessage <- function(x) {
        if(x>0)
                print("x is greater than zero")
        else
                print("x is less than or equal to zero")
        invisible(x) # Prevents R from printing the object
}

printmessage(4) # Function carries out successfully
printmessage(NA) # Message created. NA is neither greater than nor less than 0.

# Let's rewrite the previous function.
printmessage2 <- function(x) {
        if(is.na(x))
                print("x is a missing value!")
        else if(x > 0)
                print("x is greater than zero")
        else
                print("x is less than or equal to zero")
        invisible(x)
}

printmessage2(x) # Now R has ID'd the NA

```
* How do you know that something is wrong with your function?
- What was the input? How did you call the function?
- What were you expecting? Output, messages, other results?
- What did you get?
- How does what you get differ from what you were expecting?
- Were your expectations correct in the first place?
- Can you reproduce the problem exactly?
        * Remember to set seeds if you work with randomly-produced data

## Basic Debugging Tools
* traceback(): prints out the function call stack after an error occurs; does
  nothing if there is no error
* debug(): Flags a function for "debug" mode which allows you to step through
  execution of a function one line at a time
* browser(): Suspends the execution of a function wherever it is called and puts 
  the function in debug mode
* trace(): Allows you to insert debugging code into a function at specific places
* recover(): Allows you to modify the error behavior so that you can browse the
  function call stack
* print() and cat(): More blunt force









