---
title: "Week 1 Notes"
output: html_notebook
---
# Background Material
```{r}
  ## Check working directory.
  getwd()
  dir()

  ## Change working directory
  setwd("C:/Users/joshd/Desktop/grunt-work/Coursera/DataScienceJohnHopkins/02-R_Programming/")

  ## Reference function
  myfunction()

  ## Make both functions available without running it first.
  source("MyCode.R")

  ## Run second code in MyCode.R only.
  second(4)
  second(4:10)
```
* Everytime you change your code, you must use the source() function to load your code 
  functions into R again.

# Free Software
* With free software, you have the freedom to:
  - Run the program for any purpose
  - Study and adapt the program as you see fit.
  - Redistribute copies of the program.
  - Improve and release the program.

# Drawbacks of R
* No helpline; if the feature is not there, you must make it
* Limited to physical memory of machine you are on.

# Design of the R system
* The "base" R system has the base packages needed ...and everything else. Over 4000+ 
  packages

# Getting Help
* Try to find answer to something by:
  - Searching the web
  - Searching the archives of the forum you plan to post to
  - Reading the manual
  - Reading a FAQ
  - Inspection or experimentation
  - Asking a skilled friend
  - Reading the source code
* Make sure to specify that you say you have read the documentation
* Specify:
  - Steps to reproduce the problem
  - Expected output
  - What you see instead
  - The version you are using
  - Your OS
* Do:
  - Describe the Goal not the step
  - Be explicit about your question
  - Provide minumum amount of info necessary
  - Be courteous
  - Follow up with the solution if you find it
* Don't:
  - Claim you found a bug
  - Grovel as a substitute for doing your homework
  - Post homework questions on mailing lists
  - E-mail multiple mailing lists at once
  - Ask others to debug your broken code without a hint as to what sort of problem it might
    be

# R Input and Evaluation
* "<-" is the assignment operator
```{r}
  # Entering an object's name "auto-prints" its value
  x <- 1:20
  x
```

# R Data Types
  * Objects
    * 5 Atomic Classes of Objects
    - Character
    - Numeric (Real Numbers)
    - Integer
      * 1 gives you a numeric object, but 1L explicitly gives you an integer.
    - Complex (imaginary numbers)
    - Logical (T/F)
    * Most basic object is a vector, which can only contain objects of the same class
    * A LIST on the other hand, can contain objects of different classes but still be 
    represented as a vector.
  * Attributes
    - Names, dimnames
    - Dimensions
    - Class
    - Length
    - User-defined
    * attributes() lets you set attributes for an Object

# Vectors
* Done with the c() function
* When you mix objects in a vector, COERCION occurs so that every vector element is of the
  SAME CLASS!
* Explicit Coercion
  * When you want to explicitly set a class for a vector, you use the as.* functions.
```{r}
  x <- 0:6
  class(x) # Gives the class
  as.numeric(x) # Coerces to numeric
  as.logical(x) # Coerces to logical; 0 is false, everything else is TRUE
  as.character(x) # Coerces to character
  as.complex(x) # Coerces to complex
  # With a character vector
  x <- c("a", "b", "c")
  as.numeric(x) # A vector of NAs
  as.logical(x) # And again.
```

# Lists
* Can contain elements of different classes
* Double brackets demark different classes
```{r}
  x <- list(1,"a","b", T, 1 + 4i)
  x
```

# Matrices
* Matrices are vectors with a dimension attribute which can be checked with 
  dim(matrixobject)
* Matrices are constructed column-wise (lists of data fill up an entire column before 
  moving to the next)
* You can turn a vector into a matrix using the dim() function
```{r}
  m <- 1:10
  dim(m) <-c(2,5)
  m
# You can combine two vectors into a single matrix using cbind() or rbind()
  x <- 1:3
  y <- 10:12
  cbind(x,y)
  rbind(x,y)
```

# Factors
* Used to represent categorical data
* Can by ordered or unordered; an integer vector where each integer has a label
* Treated specially using modelling functions like lm() and glm()
```{r}
  x <- factor(c("yes","yes","no","yes","no"))
  x
```

* Two levels in the above factor vector
  - table(x) # Shows how many of each factor in a factor vector
  - unclass(x) # Brings the vector down to an integer vector, "n" comes before "y" in the 
    alphabet, so "no" is factor 1 and "yes" is factor 2
```{r}
  x <- factor(c("yes","yes","no","yes","no"), levels = c("yes","no")) 
  # Ordering factors manually using argument "levels"
```

# Missing Values
```{r}
  z <- c(NA,NaN, NA, NA, NaN, NaN)
  is.na(z) # Used to test objects if they are NA
  is.nan(z) # Used to test for NaNs
```
* NA values have classes also, so there are integer NAs, character NA, etc.
* A NaN value is also NA, but NA is NOT an NaN!!!

# Data Frames
* Used to store tabular data
* Represented as a special list where every element has same length
* Each element of list can be though of as a column and the length of each element is the 
  number of rows
* Data frames can store different classes of objects in each column (unlike   
  matrices/vectors)
* Data frames have special attributes, like row.names and col.names
* Created by calling read.table() or read.csv()
* Can be converted to a matrix by calling data.matrix()
```{r}
  a <- data.frame(foo = 1:4, bar = c(T,T,F,F))
  nrow(a) # Counts rows in data frame
  ncol(a) # Counts columns in data frame
```

# The Names Attribute
* Objects can have names which is useful for writing readable code and self-describing objects
```{r}
  b <- 1:3
  names(b) <- c("foo","bar","norf") # Gives name to each index in object b
```
* Lists and matrices can also have names
```{r}
c <- list(a=1,b=2,c=3)
d <- matrix(1:4,nrow=2,ncol=2)
dimnames(d) <- list(c("a","b"),c("c","d"))
```

## Summary
### Atomic Classes: Numeric, Logical, Character, Integer, Complex
### Vectors and Lists
### Factors
### Missing Values
### Data Frames
### Names

# Reading Tabular Data

## read.table() and read.csv() for reading tabular data; for writing: write.table(); arguments:
### file = name of file
### header = logical indicating header line for file
### sep = string indicating how columns are separated
### colClasses = Character vector indicating class of each column in dataset
### comment.char = character string indicating comment character
### skip = number of lines to skip from the beginning
### stringsAsFactors = which character variables should be coded as factors
### Typically figures out everything for yourself

## readLines() for reading lines of a text file; for writing: writeLines
## source() for reading in R code files; inverse of dump()
## dget for reading in R code files
## load for reading in saved workspaces
## unserialize for reading single R objects in binary form

# Reading in Larger Datasets with read.table()
## Optimize read.table() by checking out the help page for it
## If your dataset is larger than the amount of RAM on your computer, you can probably stop right there
## colClasses allows you to tell R what type of data is in each column
### if all numeric, you can just set colClasses = "numeric"
### A quick, dirty way to figure out all classes of a datatable:
initial <- read.table("datatable.txt", nrows = 100) # Only reads first 100 rows of a big table to initial
classes <- sapply(initial,class) # Reads a vector of classes to object "classes"
tabAll <- read.table("datatable.txt",colClasses=classes) # Now you can tell read.table() which classes are in which columns

## Calculating Memory Requirements
### Rows x Columns x 8 bytes/character = # Bytes
### Bytes/(2^20 Bytes/Megabyte) = # Megabytes
### MB/1000 = # Gigabytes of Memory Required
### Rule of Thumb: 2*GB = # Gigabytes ACTUALLY Required

# Textual Data Formats
## Dumping and dputing are useful because resulting format is editable
## Unlike writing out a table/csv, dump() and dput() perserve metadata
## Textual formats can work much better with version control programs like subversion or git 
##  which can only track changes meaningfully in text files
## They can also be longer-lived, with easily-identifiable corruption
e <- data.frame(a=1,b="a")
dput(e) # Makes R code to make your data frame "e" reproducible.
dput(e,file="e.R")
new.e <- dget("e.R") # dget() allows you reconstruct an object made reporducible with dput()
new.e

## Dumping Objects (dput() but for multiple objects)
f <- "foo" 
dump(c("e","f"),file="data.R") # Dumps your objects into data.R file
rm(e,f) # Removes e and f (to show you that your data is reproducible)
source("data.R") # Sources your dump and reads them back to your global environment
e # Ta
f # Da

# Interfaces to the Outside World
## Data can be read in using connection interfaces, which can be made to file or more exotic things (websites)
## Can be useful when you want to read parts of a file
g <- gzfile("words.gz") # "Opens" words.gz to to g
h <- readLines(g,10) # Reads only the first 10 lines from words.gz
## You can read lines from a webpage using url()
i <- url("https://www.jhsph.edu","r") # Connect to this site using url() and feed to object i
j <- readLines(i) # Read lines from connnection i to object j
head(j) # See the first few lines of j

# Subsetting
## Basics
### [ ] always returns object of the same class as the original; can be used to select more than one element
### [[]] is used to extract elements of a list or a data frame; it can only be used to extract a single
###  element and the class of the returned object will not necessarily be a list/data frame
### $ is used to extract element of a list/data frame by name
k <- c("a","b","c","c","d","a")
k[1] # Extracts first element in the vector k: "a"
k[1:4] # Extracts first FOUR elements in the vector k
k[k > "a"] # Subsetting with a logical index (returns only letters after "a")
l <- k > "a" # Creates logical vector
l
k[l] # Subsets according to your logical vector (F T T T T F)

## Lists
### You can subset according to an element's name, which can be useful when you don't remember where it is.
m <- list(foo = 1:4, bar = 0.6) # A main difference between a list and a vector...
m[1] # ...is indexing, where m[1] calls the entire first element "foo" (a vector), rather than just the # 1.
m[[1]] # Double bracket gives you just the object contents 
m$bar == m[["bar"]]
m$bar == m["bar"] # The single bracket returns the name of the list

### The [[]] operator can be used with COMPUTED indices; $ can only be used with LITERAL names
n <- list(foo=1:4,bar=0.6,baz="hello")
n[c(1,3)] # Calls the name and contents of the first and third elements (foo and baz) of the list
name <- "foo"
n[[name]] # Computed index for "foo"
n$name # Element "name" doesn't exist
n$foo # Element "foo", however, does exist

### You can subset nested list elements with [[]] operator
o <- list(a=list(10,12,14),b=c(3.14,2.81)) # First element in list o is list a (which has three elements itself)
o[[c(1,3)]] # Calls third element of list "a" in the first element of list "o"!
o[[c(1,3)]] == o[[1]][[3]]
o[[c(2,1)]] # Calls FIRST element of vector "b" in the second element of list "o"

## Matrices
### Can be subsetted in the usual way with (i,j) type indices
p <- matrix(1:6,2,3)
p[1,2] # Calls element in row 1 and column 2
p[2,1] # Calls element in row 2 and column 1
p[1, ] # Calls entire row 1.
p[ ,2] # Calls entire column 2; note how it is returned as a VECTOR of length 1
### By default, when a single element of a matrix is retrieved, it is returned as a vector of length 1 rather
#### than a 1x1 matrix. You can toggle this with drop=FALSE
p[1,2,drop=F] ## Returns element in row 1 and column 2 but as a 1x1 matrix!
p[1, , drop=F] ## Returns entire row 1, but as a 1x3 matrix (retaining original dimension)

## Partial Matching
### Allowed with [[]] and $
q <- list(aardvark = 1:5) # Typing out aardvark is a pain, so we can just shorten it to "a" when calling
q$a # Looks for element that starts with "a", or aardvark
q[["a"]] # Double bracket DOESN'T do partial matching, but you can pass argument exact to toggle this
q[["a", exact=F]] # It works!

## Removing missing values
### Common task is to remove missing values (NAs)
r <- c(1,2,NA,4,NA,5)
bad <- is.na(r) # Identifies all NAs in vector "r"
r[!bad] # Subset using logical vector by asking only for FALSE elements (non-NA elements)

### What there are multiple things and you want to take the subset with no missing values?
s <- c("a","b",NA,"d",NA,"f")
good <- complete.cases(r,s) # Creates logical vector of non-NA elements in both argument objects
t <- c("a","b","c","d","e","f")
good1 <- complete.cases(r,t) # Only indices that contain a non-NA element will be marked as TRUE
r[good] # Then we can call according to the complete.cases() logical vector
s[good]

### Also works for data frames
airquality # Commonwealth R file with lots of NAs
goodcases <- complete.cases(airquality) # Creates logical vector based on NA elements in matrix,
#### Each row with an NA will return a FALSE for that row's index in the logical vector
airquality[goodcases, ] # We've removed all rows with NA!

# Vectorized Operations
t <- 1:4; u <- 6:9 # Vectors interact parallely
t + u # This adds the first elements of vectors t and u, the second elements of vectors t and u, and so on.
## The benefit is that you don't need to write a loop for this kind of operation

## Matrix Operations
v <- matrix(1:4,2,2); w <- matrix(rep(10,4),2,2)
v * w # Element-wise multiplication; NOT TRUE matrix multiplication
v %*% w # THIS is true matrix multiplication.

# END OF WEEK 1